---
title: "Exploring the LOAR API"
author: "Per Møldrup-Dalum"
date: '2020-05-27'
output:
  html_document:
    df_print: paged
---

The Royal Danish Library has made public the OCR text of a large amount of newspapers published during the years from 1666 up to 1877. This newspaper collection is available at the Royal Library Open Access Repository: [LOAR](https://loar.kb.dk).

The collection can also be accessed using an API. The system under LOAR is DSPace. The API is described at [DSpace REST API](https://loar.kb.dk/rest/). In this post, I'll explore this API using R.

# Extending R

Start by loading the [Tidyverse](https://www.tidyverse.org), a library for using JSON in R, and a library for date and time.

```{r, message=FALSE}
library(tidyverse)
library(jsonlite)
library(lubridate)
```

# Finding the newspaper collection

The top level of the LOAR hierarchy is something called communities, but we don't need that concept here. Still, it it gives an idea about where the collections stems from.

```{r paged.print=FALSE}
fromJSON("https://loar.kb.dk/rest/communities") %>% select(name)
```


But, as we don't need that hierarchy level, just list the available collections.

```{r paged.print=FALSE}
fromJSON("https://loar.kb.dk/rest/collections") %>% select(name)
```

So, the newspapers are split into four collections. To get those collections, we need their ids. We'll store this list of collections and ids for later

```{r}
fromJSON("https://loar.kb.dk/rest/collections") %>%
  filter(str_detect(name, "Newspaper")) %>% 
  select(name, uuid) -> newspaper_collections
newspaper_collections
```

What can we then get from a collection? Let's look at the first using this URL

```{r}
str_c(
  "https://loar.kb.dk/rest/collections/",
  first(newspaper_collections %>% pull(uuid))
)
```


```{r}
fromJSON(str_c(
  "https://loar.kb.dk/rest/collections/",
  first(newspaper_collections %>% pull(uuid))
))
```

Which items do we have in that collection?

```{r}
str_c(
  "https://loar.kb.dk/rest/collections/",
  first(newspaper_collections %>% pull(uuid)),
  "/items"
)
```

```{r }
fromJSON(str_c(
  "https://loar.kb.dk/rest/collections/",
  first(newspaper_collections %>% pull(uuid)),
  "/items"
))
```
Let's pick the first item for a closer look

```{r}
fromJSON(str_c(
  "https://loar.kb.dk/rest/collections/",
  first(newspaper_collections %>% pull(uuid)),
  "/items"
)) %>% 
  pull(uuid) %>%
  first() -> uuid
uuid
```


```{r}
fromJSON(str_c("https://loar.kb.dk/rest/items/",uuid))
```


So, this is wierd, because even though the `bitstreams` value is `NULL`, I know it contains the actual content of the record/item. Let's look at that

```{r}
fromJSON(str_c("https://loar.kb.dk/rest/items/",uuid,"/bitstreams"))
```
And now we're close to the actual data. In the above table, the data is available in bitstream with id  `d2d3869f-ad37-461c-bcb4-79ffc7d9d0fe`, and we get it by using the `retrieve` function from the API. The content is delivered as CSV, and normally I would use the `read_csv` for such data. But this CSV format has some issues with the encoding of quotes. Therefore, we must use the more general `read_delim` function with the two `escape_` parameters.

```{r}
fromJSON(str_c("https://loar.kb.dk/rest/items/",uuid,"/bitstreams")) %>%
  filter(name == "artikler_1678.csv") %>%
  pull(retrieveLink) -> artikler_1678_link

artikler_1678_link
```



```{r}
artikler_1678 <- read_delim(
  str_c("https://loar.kb.dk/",artikler_1678_link),
  delim = ",",
  escape_backslash = TRUE,
  escape_double = FALSE)
```

```{r}
artikler_1678
```

```{r}
glimpse(artikler_1678)
```

# What can we do with the data?

To get an idea about the amount of data, let's count pages

```{r}
artikler_1678 %>% 
  group_by(sort_year_asc) %>% 
  summarise(page_count = sum(newspaper_page)) %>% 
  arrange(desc(sort_year_asc))
```
# Look at the metadata

Unfortunately, to explore the metadata, we have to download all the actual data. Hopefully this will change in the future. Still, the `bitstreams` do have a `sizeBytes` value, so let's collect those, and see how much bandwidth and storage is needed for the full collection. Well, actually for the full four newspaper collections.

So:

  for each newspaper collection
    for each item
      sum the `sizeBytes` of bitsteams with name `^artikel_`

First look at the first collection

```{r}
fromJSON(str_c("https://loar.kb.dk/rest/collections/",newspaper_collections %>% pull(uuid) %>% first() , "/items"))
```
Using that technique we can map the above used `fromJSON` function onto a list of ids, to get the items from all the newspaper collections
      
```{r}
map_df(
  newspaper_collections %>% pull(uuid),
  ~fromJSON(str_c("https://loar.kb.dk/rest/collections/", .x, "/items"))
) -> all_items
all_items
```

Now, get all the `bitstreams` associated with those items. Fist we can extract the item id

```{r}
all_items %>% pull(uuid) %>% first()
```

and from that item id, get a bitstream

```{r}
fromJSON(
  str_c(
    "https://loar.kb.dk/rest/items/",
    all_items %>% pull(uuid) %>% first(), "/bitstreams"
  )
)
```

The actual bitstream we are interested in, is the one named `artikler_1678.csv`, and we can see that it is the only one with the `CSV` format. Filter for that, and just retain the `name` and `sizeBytes`

```{r}
fromJSON(
  str_c(
    "https://loar.kb.dk/rest/items/",
    all_items %>% pull(uuid) %>% first(), "/bitstreams"
  )
) %>% 
  filter(format == "CSV") %>% 
  select(name, sizeBytes)
```

So, how do we do that for all items? Well, it should be as easy as when getting the items above

```{r, eval=FALSE}
map_df(
  all_items %>% filter(row_number() < 4) %>% pull(uuid),
  ~fromJSON(str_c("https://loar.kb.dk/rest/items/", .x, "/bitstreams"))
)
```

But I get that wierd error message?!?

Well, we have this item id: `b4fb558a-1c56-42de-8c56-7fff565bb7b4`. Which bitstreams does that give

```{r}
fromJSON(str_c("https://loar.kb.dk/rest/items/", "b4fb558a-1c56-42de-8c56-7fff565bb7b4", "/bitstreams"))
```

Okay, can we use the `map_df`for just that one item?

```{r, eval=FALSE}
c("b4fb558a-1c56-42de-8c56-7fff565bb7b4") %>% 
  map_df(
    ~fromJSON(str_c("https://loar.kb.dk/rest/items/", .x, "/bitstreams"))
  )

```

No?! Okay, what if I then select the needed columns, and here the assumption is, that one of the unneeded columns are causing the havoc.

```{r}
c("b4fb558a-1c56-42de-8c56-7fff565bb7b4") %>% 
  map_df(
    ~(fromJSON(str_c("https://loar.kb.dk/rest/items/", .x, "/bitstreams")) %>% select(name,sizeBytes))
)
```

Oh, that worked! Then try it with a few more items

```{r}
all_items %>% filter(row_number() < 4) %>% pull(uuid) %>% 
  map_df(
    ~(fromJSON(str_c("https://loar.kb.dk/rest/items/", .x, "/bitstreams")) %>% select(name,sizeBytes))
)
```

YES! Now build the data frame, that I want. First just for 3 rows (`row_number() < 4)`)
 
```{r}
all_items %>% filter(row_number() < 4) %>% pull(uuid) %>% 
  map_df(
    ~(fromJSON(str_c("https://loar.kb.dk/rest/items/", .x, "/bitstreams")) %>% select(name,sizeBytes, format))
) %>% 
  filter(format == "CSV")
```

And now with everything

```{r}
all_items %>% pull(uuid) %>% 
  map_df(
    ~(fromJSON(str_c("https://loar.kb.dk/rest/items/", .x, "/bitstreams")) %>% select(name,sizeBytes, format, uuid))
) %>% 
  filter(format == "CSV") %>% 
  select(-format) -> all_bitstreams
```

Let's have a look

```{r}
all_bitstreams
```

```{r}
summary(all_bitstreams)
```

So at last, we can get the answer to the question regarding the resources needed for downloading the complete collection. But, first lets load a library for formatting numbers in a more human readable way

```{r, message=FALSE}
library(gdata)
```

Sum all the bytes

```{r}
all_bitstreams %>% 
  summarise(total_bytes = humanReadable(sum(as.numeric(sizeBytes)), standard = "SI"))
```

That's more or less that same as intalling TeX Live ;-)

# Last example: Look at some text

Let's select a year: 1853. Let's get all the available text from that year. Now, we can cheat a bit, as we have all the bitstreams, and we can filter their names for 1853

```{r}
all_bitstreams %>% 
  filter(str_detect(name, "1853.csv"))
```

Let's get that bitstream using the `GET /bitstreams/{bitstream id}/retrieve`

```{r}
print(now())
articles_1853 <- read_delim(
  "https://loar.kb.dk/rest/bitstreams/f6543ed8-d4ba-40fe-99a8-ba26a5390924/retrieve",
  delim = ",",
  escape_backslash = TRUE,
  escape_double = FALSE)
print(now())
```

That took less that a minute, so...

Okay, what did we get

```{r}
articles_1853
```



```{r}
articles_1853 %>% 
  mutate(words = str_count(fulltext_org, boundary("word"))) %>% 
  select(-fulltext_org) %>% 
  group_by(editionId) %>% 
  summarise(total_words = sum(words)) %>% 
  arrange(desc(total_words)) -> articles_1853_word_counts
articles_1853_word_counts
```

```{r}
articles_1853_word_counts %>% 
  summarise(total = sum(total_words))
```

15 million words for one year? Well, the validity of that number is left as an exercise...

You can play with this code your self at a [public available RStudio Cloud project](https://rstudio.cloud/project/1321498). The static document is available as a RPUb at [Playing with the LOAR API](https://rpubs.com/perdalum/loar-api).

If you have any questions or comments, please reach me at pmd@kb.dk.

